---
title: "Sex hormone analysis"
output: html_notebook
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
options(max.print=200)
```
SEX HORMONE ANALYSIS IN HEALTHY FEMALE PBMCs

This script uses the pbmc dataset from the Oelen et al 2022 paper which has been analysed and annotated in the previous script - V2_pbmc.rmd

The script follows the annotation of the immune cell types, whether or not they express the sex hormone receptors such as ESR1, ESR2, GPER1, PGR, AR, FSHR, LHR, GNRHR.
The script searchers for all possible aliases of the above menstioned sex hormone receptors in the dataset and captures those that are present.
In order to not miss any genes that could have been filtered out during QC, 'all.genes' from the previous script was cross checked as well.
The script also filters out all the male samples, as the focus of the project is to understand the imapct of the sex hormones with respect to women's reproductive success.

```{r}
library(Seurat)
library(data.table)
library(dplyr)
library(ggplot2)
library(dittoSeq)
library(speckle)
library(patchwork)
library(ComplexHeatmap)
library(tidyr)
library(colorRamp2)
library(readxl)
library(scCustomize)
library(reshape2)
library(knitr)
library(tibble)
library(SingleCellExperiment)
library(viridis)
library(ggplotify)
library(lme4)
library(muscatWrapper)
library(writexl)
library(ggpubr)
library(kableExtra)
library(SeuratData)
library(limma)
library(scales)
# SET SEED FOR REPRODUCIBILITY
set.seed(123)
```
Read the rds file
```{r}
pbmc.UT <- readRDS("/home/inf-21-2023/thesis/scRNAseq_data_oelen/pbmc_UT_v2_sexannotated.rds")
all.genes <- rownames(pbmc.UT)
```
Define the sex hormone receptor genes and filter the available genes from the object
```{r}
receptor_genes <- c("^ESR1$", "^ESR2$", "^PGR$","^NR3C3$","^PR$", "^AR$", "^GPER", "^LGR1","^FSHRO","^ODG1", "^FSHR", "^LHR", "^LCGR",
"^LGR2", "^ULG5", "^LHRH","^LHRHR","LSH-R","HHG", "^GNRHR","^GRHR", "^LHRHR", "^GRH$", "^LHCGR" ,"^HHG", "^LCGR", "^LGR2", "LHR", "^ULG5", "^ODG4", "^FSH Receptor", "P23945-1", "P23945", "LH/CG-R","LSH-R","HHG", "ULG5", "LGR2", "LCGR" )

filtered_genes <- grep(paste(receptor_genes,  collapse = "|"), all.genes, value = TRUE)

cell_subtypes_order <- c(
  # T Cells (CD4 and CD8 T Cells)
  "CD4 Naive T", "Central memory CD4T", "Effector memory CD4T",
  "Naive CD8T", "Effector memory CD8T",
  # NK Cells
  "NK dim", "NK bright", "NK Proliferating",
  # Monocytes
  "CD14+ Monocytes", "CD14+ Proinflammatory Monocytes",
  "CD16+ Monocytes", "CD16+ Monocytes ISG",
  # B Cells
  "B cells", "Plasma B cells",
  # Dendritic Cells
  "cDCs", "pDCs",
  #
  "HSPCs", "Megakaryocytes"
)

receptors_order <- c("ESR1", "ESR2", "GPER1", "AR")
```
Assign the receptor expression status to positive if the expression is > 0 or negative if expression is < 0
```{r}
# Fetch the expression matrix
expr_matrix <- GetAssayData(pbmc.UT, assay = "RNA", layer = "data")

# For each receptor, assign status per cell
for (receptor in receptors_order) {
  if (receptor %in% rownames(expr_matrix)) {
    expr_values <- expr_matrix[receptor, ]
    pbmc.UT[[paste0(receptor, "_status")]] <- ifelse(expr_values > 0, "Positive", "Negative")
  } else {
    pbmc.UT[[paste0(receptor, "_status")]] <- NA
  }
}
pbmc.UT <- subset(pbmc.UT, subset = Predicted_sex == "F")

```


```{r}
# Annotate the cell types by receotor expression positive or negative
expr_data <- FetchData(pbmc.UT, vars = c("subject", filtered_genes), layer = "data")
expr_data$cell_type <- pbmc.UT$cluster_names[match(colnames(pbmc.UT), rownames(expr_data))]


# Convert the matrix to long format
expr_long <- melt(expr_data, id.vars = c("cell_type", "subject"),
                   variable.name = "gene", value.name = "expression")

# If raw counts are to be visualised just use Count = sum(expression > 0)
expr_long <- expr_long %>%
  group_by(cell_type, subject, gene) %>%
  mutate(percent_expression = sum(expression > 0) / n() * 100) %>%
  ungroup()

violin_plot_combined <- ggplot(expr_long, aes(x = factor(gene, levels = receptors_order), y = percent_expression, fill = gene)) +
  geom_violin(trim = TRUE) +  # Violin plot for Male and Female
  geom_point() +
# geom_jitter(width = 0.1, height = 0, alpha = 0.5) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 3, fill = "white") +  # Add median markers
  facet_wrap(~ cell_type, scales = "free_y") +  # Facet by cell type
  labs(x = "Sex hormone receptor", y = "Percent of cells", title = "Percentage of cells expressing the receptors") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    plot.title = element_text(size = 30, face = "bold"),
    strip.text = element_text(size = 30, face = "bold")
  ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.2)), limits = c(0,NA))+
        theme_classic2()
png("/home/inf-21-2023/thesis/scRNAseq_data_oelen/receptor_expression_levels.png", width = 4000, height = 2500, res = 300)
violin_plot_combined
dev.off()
#ggsave("/home/inf-21-2023/thesis/scRNAseq_data_oelen/receptor_expression_levels.png", violin_plot_combined, width = 20, height = 15)

```

```
```{r}
#saveRDS(pbmc.UT, file = "/home/inf-21-2023/thesis/scRNAseq_data_oelen/pbmc_UT_receptor.rds")
```
```{r}
library(report)
cite_packages()
```
```{r}
sessionInfo()

```
