---
title: "scRNAseq PBMC Oelen GSEA analysis"
Author: "Pooja"
date: "2025"
output:
  html_document
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/home/inf-21-2023/thesis/scRNAseq_data_oelen")
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
options(max.print=200)
set.seed(1919)
```
GENE SET ENRICHMENT ANALYSIS IN HEALTHY FEMALE PBMCs

This script uses the pbmc dataset from the Oelen et al 2022 paper which has been analysed and annotated in the previous script - V2_pbmc.rmd, sex hormone analysis, and the gene lists generated from the DESeq2 analysis of all the sex hormone positive vs negative cells.
The script follows the GSEA analysis of all the sex hormone receptor positive vs negative cells, and the results are visualised using dotplots and upset plots.

The BTM database is used for the GSEA analysis, and the results are filtered for significant p-values (p.adjust < 0.05).
The tsv file containing the BTM database is taken from the lab.
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(enrichplot)
library(UpSetR)
library(Seurat)
library(dplyr)
library(tidyr)
library(msigdbr)
library(patchwork)
library(ggupset)
library(readr)
library(ggtangle)
```
```{r}
BTM_plus <- read_tsv("/home/inf-21-2023/thesis/scRNAseq_data_oelen/BTM_plus_hs_t2g.tsv")

btm_db <- BTM_plus %>%
  gather(key = "term", value = "gene") %>%
  filter(!is.na(gene))
```
```{r}
receptors <- c("ESR1", "ESR2", "GPER1", "AR")
result_path <- "/home/inf-21-2023/thesis/scRNAseq_data_oelen/DE_results/"
result_df <- data.frame()

for (receptor in receptors) {
  csv_files <- list.files(result_path, pattern = paste0("_", receptor, "_DESeq2.csv$"), full.names = TRUE)

  for (file in csv_files) {
    cluster <- sub(paste0("_", receptor, "_DESeq2.csv"), "", basename(file))
    de_result <- read.csv(file, row.names = 1)
    message("Running GSEA for: ", cluster, " for receptor: ", receptor)

    if (!"avg_log2FC" %in% colnames(de_result)) next

    # Count upregulated and downregulated
    gene <- rownames(de_result)
    avglog2fc <- de_result$avg_log2FC
    p_val_adj <- de_result$p_val_adj
    p_val <- de_result$p_val

    result_df <- rbind(result_df, data.frame(cluster = cluster, receptor = receptor, gene = gene, log2fc = avglog2fc, p_val_adj = p_val_adj, p_val = p_val))
  }
    cell_types <- unique(result_df$cluster)

    make_gene_list <- function (cell_type) {
    d <- result_df %>%
        filter(cluster == !!cell_type) %>%
        dplyr::select(gene, log2fc) #ranked based on logfc

    d <- d[!duplicated(d$gene), ]

      geneList <- d$log2fc
      names(geneList) <- as.character(d$gene)
      geneList <- sort(geneList, decreasing = TRUE)
      return(geneList)
    }

    cluster_list <- lapply(cell_types, make_gene_list)
    names(cluster_list) <- cell_types
    gsea_btm <- compareCluster(cluster_list,
                            fun = "GSEA",
                            TERM2GENE = btm_db,
                            pvalueCutoff= 0.05,
                            pAdjustMethod = "BH",
                            nPermSimple = 1000000,
                            verbose = FALSE)
  dotplot_clusters <- gsea_btm  %>%
  filter(p.adjust < 0.05) %>%
  arrange(desc(NES)) %>%
  dotplot(color = "NES",
          size = "setSize",
          showCategory = 4,
          font.size = 10,
          label_format = 100)  +
  labs(x = NULL) +
  scale_fill_gradient2(low = "#0050A0",high = "#C64700", midpoint = 0, mid = "white",
							  limits = c(-max(abs(gsea_btm@compareClusterResult$NES)), max(abs(gsea_btm@compareClusterResult$NES))),
							  breaks = c(max(abs(gsea_btm@compareClusterResult$NES)),-max(abs(gsea_btm@compareClusterResult$NES))),
                              labels = c(paste0("Enriched in ", receptor, " positive"), paste0("Enriched in ", receptor, " negative")))+
                              theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
							        axis.text.y = element_text(size = 18, face = "bold"),
							        axis.title.x = element_text(size = 18, face = "bold"),
							        axis.title.y = element_text(size = 18, face = "bold"),
                                    legend.text = element_text(size = 14),
                                    legend.title = element_text(size = 16, face = "bold"))

ggsave(paste0("/home/inf-21-2023/thesis/scRNAseq_data_oelen/dotplot_btm_", receptor, ".png"), plot = dotplot_clusters)

  df <- filter(gsea_btm@compareClusterResult, p.adjust < 0.05)
  df_list <- split(df, df$Cluster)

  new_df_list <- list()

  for (i in seq_along(df_list)) {
    cluster_name <- names(df_list)[i]
    cluster_data <- df_list[[i]]
    new_df_list[[cluster_name]] <- cluster_data[["Description"]]
  }

    if (length(new_df_list) > 1) {
    upset <- upset(fromList(new_df_list), nsets = length(new_df_list), order.by = "freq", text.scale = 2.3)
    ggsave(paste0("/home/inf-21-2023/thesis/scRNAseq_data_oelen/upset_plot_", receptor, ".png"), ggplotify::as.ggplot(upset))
    }
  }
}
```

```{r}
library(report)
cite_packages()
```
```{r}
sessionInfo()

```